A Data-Driven Early Warning System for Mining Accident

### 2017-03-21 San Antonio/slides/presentation_slides.Rmd
updated slides

- Process MSHA safety data
- Understand the underlying causal relationships
- Develop early warning systems

@@

## Primary Datasets

@@

@@

- 664,128 rows
- 10,377 unique mines
- From 2000 to 2015 in quarters
- Each row represents data for a unique combination of mine, year, and quarter
    - e.g., Upper Big Branch Mine in the second quarter of 2010
- Each row contains both current and past information

@@

## Violation Trend: Crandall Canyon

@@

```{r gibson}
if(!exists("complete.active.quarters")){load("./San Antonio/output/result.RData")}
yearly.viol <- complete.active.quarters %>% 
  filter(mine_id == 1202215 & year < 2007) %>%
  group_by(year) %>%
  summarise(count = sum(viol.quantity))
qplot(x = year, y = count, data = yearly.viol, geom = c("point", "line"),
      main = "Number of Yearly Violations",
      xlab = "Year", ylab = "Count")
```

@@

```{r severe all}
if(!exists("complete.active.quarters")){load("./San Antonio/output/result.RData")}
severe.summary <- complete.active.quarters %>%
  filter() %>%
  mutate(severe = ifelse(
    num.death + num.dis > 0, TRUE, FALSE
  )) %>%
  group_by(severe) %>%
  summarise(n = n()) %>%
  mutate(perc = n / sum(n) * 100)
print(severe.summary)
```

@@

## Processing Data

- Remove **inactive** quarters

<!-- http://www.stata.com/manuals13/rclogit.pdf -->
## Processing Data

- Consider fixed-mine effects
    - Biostatisticians and epidemiologists call it "conditional logistic regression" (R, survival::clogit)
    - Suitable for panel data (e.g., our master data table)
    - Model includes mine-specific but time-invariant variables (e.g., same slope but different intercepts for different mines)

@@

- Sort result based on number of deaths

```{r glm rank death}
if (!exists("glm.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(glm.result %>% filter(year >= 2005) %>% arrange(desc(num.death), mine.name), 10) %>% select(
  mine.name, num.death, severe, prob, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```

<!--
## Logistic Regression Without Fixed Effects

- Sort result based on model probability

```{r glm rank prob}
if (!exists("glm.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(glm.result %>% filter(year >= 2005) %>% arrange(desc(prob)), 10) %>% select(
  mine.name, num.death, severe, prob, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```
-->

@@

- Sort result based on number of deaths

```{r fe.div rank death}
if (!exists("glm.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(fe.div.result %>% filter(year >= 2005) %>% arrange(desc(num.death), mine.name), 10) %>% select(
  mine.name, num.death, severe, prob, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```

@@

- Sort result based on model probability

```{r fe.div rank prob}
if (!exists("glm.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(fe.div.result %>% filter(year >= 2005) %>% arrange(desc(prob)), 10) %>% select(
  mine.name, num.death, severe, prob, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```

@@

```{r top 10 days lost}
if(!exists("complete.active.quarters")){load("./San Antonio/output/result.RData")}
top.days.lost <- complete.active.quarters %>% filter(active) %>% arrange(desc(num.days.lost), mine.name) %>% select(mine.name, year, quarter, num.days.lost)
head(top.days.lost, 10)
```

@@

```{r new era viol}
if(!exists("complete.active.quarters")){load("./San Antonio/output/result.RData")}
yearly.viol <- complete.active.quarters %>% 
  filter(mine_id == 1102752 & year < 2006) %>%
  group_by(year) %>%
  summarise(count = sum(viol.quantity))
qplot(x = year, y = count, data = yearly.viol, geom = c("point", "line"),
      main = "Number of Yearly Violations",
      xlab = "Year", ylab = "Count")
```

## Simple Linear Model on Number of Days Lost

@@

<!-- - Causal relationship between accidents and historical data -->
- TBD

>>>>

- Process MSHA safety data
- Understand the underlying causal relationships
- Develop early warning systems based on past behaviors
- Credit rating/score analogy
    - Predict default probability within 18 months
    - Accidents: defaults a month or a year prior to application
    - Violations: missed payments, late payments, etc.

@@

## Department of Labor Enforcement Data

- Link: <https://enforcedata.dol.gov/views/data_catalogs.php>
- Updated daily or weekly
- Enforecement Data Catalog
    - MSHA Enforcement Data
    - OSHA Enforcement Data
    - Wage and Hour Compliance Action Data
    - OFCCP Compliance Evaluation and Complaint Investigation Data
    - EBSA Enforcement Data

## MSHA Data: Sources

@@

## MSHA Data: Advantages

- Each mine has a unique mine ID, e.g., Upper Big Branch (4608436)
- Rich details, e.g., selected attributes from the accidents table:

```{r msha accidents}
if(!exists("accidents")) {load("./San Antonio/data/secret_ingredients.RData")}
names(accidents[c(1, 14, 5:13, 19:20, 37:59)])
```

## MSHA Data: Challenges

- Missing data
- Inactive mines
- Most data are not numeric
- Relatively few severe accidents ($\sim0.5\%$)

## Model Preliminaries

- Create a master data table with selected attributes
- Plot violation trends for Upper Big Branch Mine and Sago Mine
- Design a predictive model

@@

- Combine and summarize accidents/violations based on mines
- 664,128 rows, 10,377 unique mines
- From 2000 to 2015 in quarters
- Each row represents data for a unique combination of mine, year, and quarter
    - e.g., Upper Big Branch Mine in the second quarter of 2010
- Each row contains both current and past information

@@

<!--
## Violation Trend: Crandall Canyon

@@

```{r gibson}
if(!exists("complete.active.quarters")){load("./San Antonio/output/result.RData")}
yearly.viol <- complete.active.quarters %>% 
  filter(mine_id == 1202215 & year < 2007) %>%
  group_by(year) %>%
  summarise(count = sum(viol.quantity))
qplot(x = year, y = count, data = yearly.viol, geom = c("point", "line"),
      main = "Number of Yearly Violations",
      xlab = "Year", ylab = "Count")
```
-->

@@

<!-- http://www.stata.com/manuals13/rclogit.pdf -->
## Fixed-Mine Effects

- Biostatisticians and epidemiologists call it "conditional logistic regression" (`survival::clogit`)
- Suitable for panel data (e.g., our master data table)
- Model includes mine-specific but time-invariant variables (e.g., same slope but different intercepts for different mines)
- Logistic regression (for every mine)

\[\Pr(Y=1|\mathbf{X})=F(\mathbf{X}) = \frac{1}{1+e^{-(\alpha+\boldsymbol\beta\mathbf{X})}}\]

- Logistic regression with fixed effects (for the $i$-th mine)

\[\Pr(Y=1|\mathbf{X},i)=F(\mathbf{x},i) = \frac{1}{1+e^{-(\alpha_i+\boldsymbol\beta\mathbf{x})}}\]

@@

@@

- True positives

```{r glm rank death}
if (!exists("glm.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(glm.result %>% filter(year >= 2005) %>% arrange(desc(num.death), mine.name), 10) %>% select(
  mine.name, year, quarter, severe, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```

- Top predicted positives (possibly false positives)

```{r glm rank prob}
if (!exists("glm.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(glm.result %>% filter(year >= 2005) %>% arrange(desc(prob), mine.name), 10) %>% select(
  mine.name, year, quarter, severe, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```

@@

- Top true positives (deadliest accidents)

```{r fe.div rank death}
if (!exists("fe.div.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(fe.div.result %>% filter(year >= 2005) %>% arrange(desc(num.death), mine.name), 10) %>% select(
  mine.name, year, quarter, severe, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```

@@

- Top predicted positives (possibly false positives)

```{r fe.div rank prob}
if (!exists("fe.div.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(fe.div.result %>% filter(year >= 2005) %>% arrange(desc(prob)), 10) %>% select(
  mine.name, year, quarter, severe, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```

<!--
## New Era Mine

@@

```{r top 10 days lost}
if(!exists("complete.active.quarters")){load("./San Antonio/output/result.RData")}
top.days.lost <- complete.active.quarters %>% filter(active) %>% arrange(desc(num.days.lost), mine.name) %>% select(mine.name, year, quarter, num.days.lost)
head(top.days.lost, 10)
```
-->

@@

```{r new era viol}
if(!exists("complete.active.quarters")){load("./San Antonio/output/result.RData")}
yearly.viol <- complete.active.quarters %>% 
  filter(mine_id == 1102752) %>%
  group_by(year) %>%
  summarise(count = sum(viol.quantity))
qplot(x = year, y = count, data = yearly.viol, geom = c("point", "line"),
      main = "Number of Yearly Violations",
      xlab = "Year", ylab = "Count")
```

## New Labels Including Days Lost

- Previously defined severe accidents and days lost $>$ 300

```{r fe.div2}
if (!exists("fe.div2.result")){load("./San Antonio/output/results_fixed_effects.RData")}
print(fe.div2.performance$table)
print(fe.div2.performance$byClass)
```

## New Labels Including Days Lost

- Top true positives (deadliest accidents)

```{r fe.div2 rank death}
if (!exists("fe.div2.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(fe.div2.result %>% filter(year >= 2005) %>% arrange(desc(num.death), mine.name), 10) %>% select(
  mine.name, year, quarter, severe, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```

## New Labels Including Days Lost

- Top predicted positives (possibly false positives)

```{r fe.div2 rank prob}
if (!exists("fe.div2.result")){load("./San Antonio/output/results_fixed_effects.RData")}
head(fe.div2.result %>% filter(year >= 2005) %>% arrange(desc(prob)), 10) %>% select(
  mine.name, year, quarter, severe, pred
) %>% mutate (mine.name = strtrim(mine.name, 40))
```

<!-- - Causal relationship between accidents and historical data -->
- Summary
    - Two deadliest mine accidents in the last decade: Upper Big Branch & Sago
    - MSHA

## Appendix: Simple Linear Model

@@
